State of the Practice for Lattice Boltzmann Method Software

### 2022-03-04 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Complete full draft of SOP for LBM paper

    We analyze the state of software development practice for Lattice Boltzmann
    solvers by quantitatively and qualitatively measuring and comparing 24
    software packages for 10 software qualities (installability, correctness/
    verifiability, reliability, robustness, usability, maintainability,
    reusability, understandability, visibility/transparency and
    reproducibility). Our reproducible analysis method employs a measurement
    template (containing 103 measures that are manually and automatically
    extracted from the project repository) and developer interviews (consisting
    of a set of 20 questions). From the measurement results, we ranked the
    software using the Analytic Hierarchy Process (AHP). Our ranking was roughly
    consistent with ranking the repositories by the number of GitHub stars,
    although the number one project by star count (Sailfish) is ranked 9th by
    our methodology. Our top three packages were ESPResSo, Ludwig, and Palabos
    (all in the top 10 by star count). We interviewed 4 developers to gain
    insight into their current pain points. Identified challenges include lack
    of development time, lack of funding, and difficulty with ensuring
    correctness.  Based on our assessment, we make the following recommendations
    for current and future LBM projects: provide a detailed user manual and
    tutorials, explicitly state the limits of the software, use user-friendly
    programming languages, consider a peer review process, communicate
    development and contribution information, and use continuous integration and
    project management tools.  \wss{Change the abstract to discuss how LBM
    compares to common practices and best practices. Highlight some future
    practices.  \wss{Follow the general research software trend of increasing
    use of version control systems and project management on GitHub or
    similar.}}

@@

To potentially improve the quality of LBM software, and help future users select
the right package, our goal was to analyze the current state of LBM software
development, including determining how well current best practices are followed
and finding pain points for developers.  We compared the state of the practice
for LBM software to the development practices recommended and employed for
research software in general. To improve the quality of LBM software we looked
for examples where developers have been successful so that we can share these
with the LBM community, and with the broader research software community.  We
also aimed to identify potential approaches for mitigating the identified pain
points in the future.

Our methodology involved the following steps: i) Identify a list of 45 candidate
software packages; ii) Filter the list to a length of 24 packages; iii) Collect
repository related data on each software package, like number of stars, number
of open issues, number of lines of code; iv) Fill in the measurement template
(the template consists of 108 questions to assess 9 qualities (installability,
correctness/verifiability, reliability, robustness, usability, maintainability,
reusability, understandability, and visibility/transparency)   The repository
data is collected by filling in a 108 question measurement template, which
requires installing the software on a VM, running simple tests (like completing
the getting started instructions (if present)), and searching the code,
documentation and test files.; v) Rank the software using the Analytic Hierarchy
Process (AHP); vi) Interview developers (the interview consists of 20 questions
and takes about an hour); and, vii) Conduct a domain analysis. The collected
data is analyzed by: a) comparing the ranking by best practices against the
ranking by popularity; b) comparing artifacts, tools and processes to current
research software development guidelines; and, c) exploring pain points.

Table~\ref{topperformerstable} summarizes the top 5 performers for each of the qualities that we assessed.

@@

Comparison to community.

Comparison of artifacts.

Comparison of tools.

Comparison of processes.

Developer pain Points

Lessons from developers

Recommendations, starting from current best practices.

\begin{enumerate}
	\item \textbf{Follow the example of LBM packages that already follow best practices.}  Point to the table summarizing the winners for each quality.  Review the practices from the previous section.  Addresses PP?.
	\item \textbf{Continuous integration.} Addresses PP?.
	\item \textbf{Document API.} Addresses PP?.
\end{enumerate}

Future work

\wss{More on the lessons to learn from the top packages.  The comparison to
common practice - is LBM ahead or behind?  More on best practices comparison.
Future practices.}

The goal of this assessment isn't to rank the software, but to use the ranking
exercise as a means to understand the state of the practice of LBM software
development.

We understand that software packages vary in goals, developers, funding, and
other aspects. Our goal was to highlight software quality successes through an
evaluation of the entire domain software family. Furthermore, threats to the
validity of the findings are highlighted in Section~\ref{threats}.

Our comparison may point out areas where some LBM software packages fall short
of current best practices.  This is not intended to be a criticism of any
existing packages, especially since in practice not every project needs to
achieve the highest possible quality.  However, rather than delve into the
nuances of which software can justify compromising which practice we will write
our comparison under the ideal assumption that every project can have sufficient
resources to match best practices.

Many software packages would benefit from a rewrite or reorganization of
installation instructions. A single location for installation instructions would
improve their maintainability and correctness. Listing compatible operating
system and dependency versions would decrease installation time and errors, as
would adding instructions on installing dependencies. Installation process
errors should prompt the system to display detailed messages. Once a software
package is installed, either an automatic validation needs to be performed or
the user needs to be able to perform a manual validation using test examples
that include expected output. Finally, uninstallation instructions should be
included in the documentation. \wss{Have to decide how much to emphasizse
installability.}  \wss{Maybe at some point should mention more modern
installation support, since most of the projects use make, and there are other
options.}

Ensuring software quality can take significant time. An inability to develop
high quality documentation due to time constraints could be mitigated in the
future by the use of automatic document generation tools like Drasil. The Drasil
Framework consists of a collection of Domain Specific Languages (DSL) for
capturing scientific documents, structures, and computing knowledge, and then
transforming this knowledge into relevant software artifacts without having to
manually duplicate knowledge into multiple artifacts \citep{SzymczakEtAl2016}.

Future State Of The Practice Assessments

As mentioned in Section~\ref{threats}, the measures listed in our measurement
template may not be broad enough to accurately capture some qualities. Adding or
extending measures is worth considering. For example, usability experiments and
performance benchmarks could be incorporated into the assessment.

Section~\ref{SecReproducibility} noted that developers were not asked to comment
on the reproducibility of their source code from their requirements
specifications and design documentation. Adding this question to the interview
guide should be considered in the next iteration of state of the practice
assessments. Furthermore, as mentioned in Section~\ref{threats}, it was found
that some pertinent information was not specified in public artifacts. The use
of unit testing and continuous integration by several packages (OpenLB, pyLBM,
TCLB) was only discovered during interviews with developers. Adding further
questions to the interview guide regarding the measures that are on the
measurement template could reduce instances of incorrect data being collected.
This additional interview data could be analyzed and incorporated into the AHP
ranking, ensuring quality designations more accurate represent the true quality
of the software packages.  \wss{clarify the distinction between usability and
understandability.}

For the current project, we did not assess the stronger form of
``reproducibility'' known as replicability.  A software project is considered
replicable if the documentation alone (without the code) is sufficiently precise
and complete to allow a third-part to re-obtain the results of the original code
\citep{BenureauAndRougier2017}.  Replicability is difficult to achieve with
current documentation practices.  This question should be considered in the next
iteration of state of the practice assessments.

The final result of the domain analysis will be tables of commonalities,
variabilities, and parameters of variation of a program family.
\cite{smith2008commonality} present a template for conducting a commonality
analysis, which was referred to when conducting this work.
\cite{weiss1998commonality} describes another commonality analysis technique for
deciding the members of a program family. \cite{SmithAndChen2004} and
\cite{SmithMcCutchanAndCarette2017} are examples of a commonality analysis for a
family of mesh generating software and a family of material models,
respectively. The steps to produce a commonality analysis are:

@@

We would like to thank Ao Dong and Oluwaseun Owojaiye for many fruitful
discussions on topics relevant to this paper.

>>>>

    We analyze the state of software development practice for Lattice Boltzmann
    solvers and find by quantitatively and qualitatively measuring and comparing
    24 software packages for 10 software qualities (installability, correctness/
    verifiability, reliability, robustness, usability, maintainability,
    reusability, understandability, visibility/transparency and
    reproducibility). Our reproducible analysis method employs a measurement
    template (containing 108 measures that are manually and automatically
    extracted from the project repository) and developer interviews (consisting
    of a set of 20 questions). From the measurement results, we ranked the
    software using the Analytic Hierarchy Process (AHP). Our ranking was roughly
    consistent with ranking the repositories by the number of GitHub stars,
    although the number one project by star count (Sailfish) is ranked 9th by
    our methodology. Our finding is that the state of the practice is healthy
    with 67\% of the measured packages ranking in the top 5 for at least one
    quality, the majority of LBM generated artifacts corresponding to general
    recommendations from research software developers, common utilization (67\%
    of packages) of version control and adoption of a quasi-agile development
    process.Areas of best practice to potentially improve include adoption of
    continuous integration, API documentation and enforcement of programming
    style guides.  We interviewed 4 developers to gain insight into their
    current pain points. Identified challenges include lack of development time,
    lack of funding, and difficulty with ensuring correctness.  Developers are
    addressing these pain points by designing for change, circumventing the
    oracle problem and prioritizing documentation and usability.  For future the
    best practices of the future, we suggest the community consider: employing
    linters, conducing rigorous peer reviews, writing and submitting more papers
    on software, growing the number of contributors and augmenting the theory
    manuals to include more requirements relevant information.

@@

To help users select the right LBM package for their needs, and to potentially
provide direction for future LBM software development, our goal was to analyze
the current state of LBM software development.  We aimed to determine how well
best practices are followed and to understand current developer pain points.  We
compared the state of the practice for LBM software to the development practices
recommended and employed for research software in general. To improve the
quality of LBM software we looked for examples where developers have been
successful so that we can share these with the LBM community, and with the
broader research software community.  We also aimed to identify potential
approaches for mitigating the identified pain points in the future.

Our analysis points to areas where some LBM software packages fall short of
current best practices.  This is not intended to be a criticism of any existing
packages, especially since in practice not every project needs to achieve the
highest possible quality.  However, rather than delve into the nuances of which
software can justify compromising which practice we conducted our comparison
under the ideal assumption that every project can have sufficient resources to
match best practices.

Our methodology involved the following steps: i) Identify a list of 45 candidate
software packages; ii) Filter the list to a length of 24 packages; iii) Collect
repository related data on each software package, like number of stars, number
of open issues, number of lines of code; iv) Fill in the measurement template
(the template consists of 108 questions to assess 9 qualities (installability,
correctness/verifiability, reliability, robustness, usability, maintainability,
reusability, understandability, and visibility/transparency).  Filling in the
template requires installing the software on a VM, running simple tests (like
completing the getting started instructions (if present)), and searching the
code, documentation and test files; v) Rank the software using the Analytic
Hierarchy Process (AHP); vi) Interview developers (the interview consists of 20
questions and takes about an hour); and, vii) Conduct a domain analysis. The
collected data was analyzed by: a) comparing the ranking by best practices
against the ranking by popularity; b) comparing artifacts, tools and processes
to current research software development guidelines; and, c) exploring pain
points.  Our ranking was roughly consistent with ranking the repositories by the
number of GitHub stars, although the number one project by star count (Sailfish)
is ranked 9th by our methodology.

Current LBM software provides many examples of following best practices, to the
benefit of users and developers.  This is illustrated by
Table~\ref{topperformerstable}, which summarizes the top 5 performers for each of
the identified qualities and for the overall quality. Taking the union of all
entries in the table, we have 16 top performers: ESPREsSo, MechSys, OpenLB,
DL\_MESO, Sailfish, Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce,
ESPREsSo++, LUMA, MP-LABS, Palabos, and HemeLB.  Therefore, two thirds (67\%) of
the 24 measured packages appear in the top 5 for at least one quality.  Of these
entries almost all of them are classified as alive, except for three packages:
laboetie, MP-LABS, and HemeLB.  This suggests that alive packages are in a
healthy state with respect to adopting best practices.

@@

Further evidence of the healthy state of the practice can be seen by comparing
the observed software artifacts versus those recommended by software engineering
guidelines.  As Section~\ref{Sec_CompareArtifacts} shows, the majority of LBM
generated artifacts correspond to general recommendations from research software
developers.  LBM software even seems to be ahead of best practices with an
example of a video user guide for one of the packages.  Comparing LBM with
respect to tool usage (Section~\ref{Sec_CompareTools}) shows that LBM is keeping
pace with other research software, with 67\% of the surveyed tools using version
control.  The state of the practice for the software development process of LBM
software matches the quasi-agile process generally adopted for research software
(Section~\ref{Sec_CompareMethodologies}).  The best practice process of peer
review is employed by some LBM developers.

Although the state of the practice for LBM software is healthy, we have noted
areas for potential improvement in the adoption of best practices, as follows:

\begin{itemize}
	\item Increase the utilization of continuous integration.
	\item Increase the amount of API documentation.
	\item Add roadmap documentation showing what is planned for the future.
	\item Add a code of conduct to explicitly show how developers should treat
	one another.
	\item Increase the use and enforcement of programming style guides.
	\item Use checklists to ensure best practices are followed by developers.
	\item Include uninstall instructions.
	\item Explicitly document the software's motivation and design principles.
\end{itemize}

To advance the state of the practice beyond current best practices toward future
best practices, we interviewed 4 developers to learn about their pain points
(Section~\ref{painpoints}).  The identified pain points include lack of
development time, lack of software development experience, lack of incentive and
funding, lack of external support, technology hurdles, ensuring correctness,
usability, technical debt and documentation quality.  We found
(Section~\ref{Sec_AddressConcerns} that developers are currently creating
practices to address these pain points, including: designing for change,
circumventing the oracle testing problem, and prioritizing documentation and
usability.  To further plan for future best practices, we identified eight
potential directions (Section~\ref{Sec_Recommendations}), in order of increasing
novelty:

\begin{enumerate}
	\item Employ linters.
	\item Conduct rigorous peer reviews.
	\item Write and submit papers on software.
	\item Grow the number of contributors.
	\item Further address the oracle problem for testing.
	\item Augment theory manuals to include requirements information.
	\item Improve re-runnability, repeatability, reproducibility and
	replicability.
	\item Generate all things.
\end{enumerate}

Future work can be done to address the threats to validity mentioned in
Section~\ref{threats}.  For instance, the measures listed in our measurement
template could be broadened to better capture some qualities. For example,
usability experiments and performance benchmarks could be incorporated into the
assessment.  Section~\ref{threats} also mentions that some pertinent information
on continuous integration and unit testing was not specified in public
artifacts. Adding further questions to the interview guide regarding the
measures that are on the measurement template could reduce instances of
incomplete data. This additional interview data could be analyzed and
incorporated into the AHP ranking, ensuring quality designations more accurate
represent the true quality of the software packages. In the future we could also
expand the interview questions to go beyond reproducibility to also discuss
replicability. Furthermore, in future interviews we would like to clarify the
distinction between usability and understandability.

Our domain analysis (Section~\ref{SecDomainAnalysis} was relatively shallow.  In
the future we would like to dig more deeply into this analysis.  We would like
to produce tables of commonalities, variabilities, and parameters of variation
for the family of LBM solvers. We could follow the template from
\citep{smith2008commonality}, or adopt the analysis techniques from
\citep{weiss1998commonality}. \cite{SmithAndChen2004} and
\cite{SmithMcCutchanAndCarette2017} are examples of a commonality analysis for a
family of mesh generating software and a family of material models,
respectively.

@@

We would like to thank Ao Dong, Oluwaseun Owojaiye, Mohammad Ali Daeian and Reza
Sadeghi for fruitful discussions on topics relevant to this paper.
